---
title: "brnet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{brnet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(mcbrnet)
library(tidyverse)
```

## Basic usage

`brnet()` generates a random branching network with the specified number of patches and the probability of branching. The key arguments are the number of habitat patches (`n_patch`) and the probability of branching (`p_branch`), which users must specify. With these parameters, the function generates a branching network through the following steps:

1.  Draw the number of branches in the network. An individual branch is defined as a series of connected patches from one confluence (or outlet) to the next confluence upstream (or upstream terminal). The number of branches in a network $N_{b}$ is drawn from a binomial distribution as $N_{b} \sim Binomial(N,P_{b})$, where $N$ is the number of patches and $P_b$ is the branching probability.

2.  Draw the number of patches in each branch. The number of patches in branch $q$, $n_q$, is drawn from a geometric distribution as $n_{q} \sim Ge(P_b)$ but conditional on $\sum_{q}^{N_{b}} n_q = N$.

3.  Organize branches into a bifurcating branching network.

The function returns:

-   `adjacency_matrix`: adjacency matrix.

-   `distance_matrix`: distance matrix. Distance between patches is measured as the number of steps required to reach from the focal patch to the target patch through the network.

-   `df_patch`: data frame (`dplyr::tibble`) containing patch attributes.

    -   *patch_id*: patch ID.
    -   *branch_id*: branch ID.
    -   *environment*: environmental value at each patch (see below for details)
    -   *disturbance*: disturbance level (i.e., proportional mortality) at each patch (see below for details)
    -   *n_patch_upstream*: the number of upstream contributing patches (including the focal patch itself; akin to the watershed area in river networks).

## Quick start

The following script produce a branching network with `n_patch = 50` and `p_branch = 0.5`. By default, `brnet` visualizes the generated network using functions in R packages `igraph` and `ggraph` (`plot = FALSE` to disable):

```{r, include=F}
set.seed(1)
```

```{r}
net <- brnet(n_patch = 50, p_branch = 0.5)
```

By default, patches are colored based on environmental values.

To view matrices, type the following script:

```{r}
# adjacency matrix
# showing 5 patches for example
net$adjacency_matrix[1:5, 1:5]

```

```{r}
# distance matrix
# showing 5 patches for example
net$distance_matrix[1:5, 1:5]

```

The following script shows patch ID, branch ID, environmental values, disturbance values, and the number of upstream contributing patches for each patch:

```{r}
net$df_patch
```

## Customize: visualization

**Arguments:** `patch_label`, `patch_size`

Users may add patch labels using the argument `patch_label`:

```{r, include=F}
set.seed(1)
```

```{r viz1}
# patch ID
brnet(n_patch = 50,
      p_branch = 0.5) %>% 
  ggbrnet(patch_label = "patch")
```

```{r, include=F}
set.seed(1)
```

```{r viz2}
# branch ID
brnet(n_patch = 50,
      p_branch = 0.5) %>% 
  ggbrnet(patch_label = "branch")
```

```{r, include=F}
set.seed(1)
```

```{r viz3}
# number of upstream contributing patches
brnet(n_patch = 50,
      p_branch = 0.5) %>% 
  ggbrnet(patch_label = "n_upstream")
```

To change patch size, specify `patch_size`:

```{r, include=F}
set.seed(1)
```

```{r viz4}
brnet(n_patch = 50,
      p_branch = 0.5) %>% 
  ggbrnet(patch_size = 1)
```

## Customize: environment

**Arguments:** `mean_env_source`, `sd_env_source`, `rho`, `sd_env_lon`

Environmental values are generated as detailed below:

1.  Environmental values for upstream terminal patches are drawn from a normal distribution as $z_h \sim N(\mu_{h}, \sigma_h^2)$ (arguments `mean_env_source` and `sd_env_source`).

2.  Downstream environmental values are determined by an autoregressive process as $z_{down} \sim N(\rho z_{up}, \sigma_l^2)$ (argument `sd_env_lon`). At bifurcation patches (or confluence), the environmental value takes a weighted mean of the two contributing patches given the size of these patches $N_{up}$ (the number of upstream contributing patches): $z_{down} = \omega(\rho z_{up, 1} + \epsilon_1) + (1 - \omega)(\rho z_{up,2} + \epsilon_2)$, where $\omega = \frac{N_{up,1}}{N_{up,1} + N_{up,2}}$ and $\epsilon \sim N(0, \sigma_l^2)$.

Users may change the values of $\mu_h$ (default: `mean_env_source = 0`), $\sigma_h$ (`sd_env_source = 1`), $\rho$ (`rho = 1`), and $\sigma_l$ (`sd_env_lon = 0.1`). Increasing the value of `sd_env_source` leads to greater variation in environmental values at upstream terminals. The argument `rho` determines the strength of longitudinal autocorrelation (the greater the stronger autocorrelation). The argument `sd_env_lon` determines the strength of longitudinal environmental noise.

```{r, include = F}
set.seed(1)
```

```{r env1, eval = T}
# with large variation at headwaters
brnet(n_patch = 50,
      p_branch = 0.5,
      sd_env_source = 3,
      sd_env_lon = 0.5,
      rho = 0.5) %>% 
  ggbrnet()
```

```{r env2, eval = T}
# large local variation with no headwater variation
brnet(n_patch = 50,
      p_branch = 0.5,
      sd_env_source = 0,
      sd_env_lon = 3,
      rho = 0.5) %>% 
  ggbrnet()
```

## Customize: disturbance

**Arguments:** `mean_disturb_source`, `sd_disturb_source` , `sd_disturb_lon`

Disturbance values are generated as detailed below:

1.  Disturbance levels for upstream terminal patches (i.e., patches with no upstream patch) are drawn from a normal distribution in a logit scale as $logit~m_h \sim N(logit~\mu_{m}, \sigma_{h,m}^2)$.

2.  Disturbance cascades downstream as in environmental values: $logit~m_{down} \sim N(logit~m_{up}, \sigma_{l,m}^2)$. At bifurcation patches (or confluence), the disturbance value takes a weighted mean of the two contributing patches given the stream size of these patches $N_{up}$ (the number of upstream contributing patches): $logit~m_{down} = \omega(logit~m_{up, 1} + \epsilon_1) + (1 - \omega)(logit~ m_{up,2} + \epsilon_2)$, where $\omega = \frac{N_{up,1}}{N_{up,1} + N_{up,2}}$ and $\epsilon \sim N(0, \sigma_{l,m}^2)$.

Users may change the values of $\mu_{m}$ (`mean_disturb_source`), $\sigma_{h,m}$ (`sd_disturb_source`) and $\sigma_{l,m}$(`sd_disturb_lon`).

```{r disturb1, eval = T}
# with large variation at headwaters
brnet(n_patch = 50,
      p_branch = 0.5,
      sd_disturb_source = 3,
      sd_disturb_lon = 0.5) %>% 
  ggbrnet(patch_color = "disturb")
```

```{r disturb2}
# with large local variation with no headwater variation
brnet(n_patch = 50,
      p_branch = 0.5,
      sd_disturb_source = 0,
      sd_disturb_lon = 3) %>% 
  ggbrnet(patch_color = "disturb")
```

## Customize: environmental pollutants

**Arguments:** `x`, `n_source`, `p`, `q`, `pattern`

From v.1.3.0, mcbrnet offers `ptsource()` function to simulate propagation of environmental pollutants in a branching network. `ptsource()` can specify:

-   `x` `brnet()` object

-   `n_source` number of point sources

-   `p` strength of downstream propagation

-   `q` strength of upstream propagation

-   `pattern` spatial pattern of point sources - `"random"`, `"cluster"`, `"upstream"`, `"downstream"`

The function takes `brnet()` object as the first argument, so it's compatible with pipe `%>%`. The concentration of pollutants has a value of 1.0 at point sources and decays with distance as $p^d$ in downstream ($d=$ distance, $p=$ geometric coefficient of distance decay) and $q^d$ in upstream. For example, if a focal location is 3 patches downstream from the point source with $p = 0.8$, the concentration would be $0.8^3 = 0.512$. By setting different values of `p` and `q`, users can model the asymmetric propagation of environmental pollutants.

For example, the following script will add a new column `impact` that represents the concentration of hypothetical environmental pollutant:

```{r pollute1}
y <- brnet(n_patch = 10) %>% 
  ptsource(n_source = 3,
           p = 0.8,
           q = 0) 

y$df_patch
```

Also, easy to visualize the spatial pattern of pollution with `ggbrnet()`. Just need to set `patch_color = "other"` and `value_col = "impact"`.

```{r pollute2}
brnet(n_patch = 30) %>% 
  ptsource(n_source = 3,
           p = 0.8,
           q = 0) %>% 
  ggbrnet(patch_color = "other", 
          value_col = "impact")
```

Finally, it is possible to specify the spatial pattern of point sources. By default, the function generates point sources at random, but it may be more realistic to have some spatial clusters.

The `"cluster"` pattern randomly pick one location as a point source, and the rest of point sources appear close to this initial point source:

```{r pollute3}
brnet(n_patch = 30) %>% 
  ptsource(n_source = 3,
           p = 0.8,
           q = 0,
           pattern = "cluster") %>% 
  ggbrnet(patch_color = "other", 
          value_col = "impact")
```

The `"upstream"` pattern pick point source locations from smaller tributaries (i.e., based on number of upstream patches), but not necessarily clustered:

```{r pollute4}
brnet(n_patch = 30) %>% 
  ptsource(n_source = 3,
           p = 0.8,
           q = 0,
           pattern = "upstream") %>% 
  ggbrnet(patch_color = "other", 
          value_col = "impact")
```

The `"downstream"` does the opposite (pick from larger streams):

```{r pollute5}
brnet(n_patch = 30) %>% 
  ptsource(n_source = 3,
           p = 0.8,
           q = 0,
           pattern = "downstream") %>% 
  ggbrnet(patch_color = "other", 
          value_col = "impact")
```

## Customize: fragmentation

**Arguments:** `x`, `rate`, `pattern`, `p`, `n_barrier`

`frgm()` will impose fragmentation to a network. The function can specify:

-   `x` 'brnet' object or adjacency matrix

-   `rate` rate parameter of exponential dispersal kernel (i.e., species attribute). This parameter defines the result dispersal matrix.

-   `pattern` fragmentation pattern; either "random" (random location of fragmentation), "upstream" (upstream edges are more likely to be fragmented), or "downstream" (downstream edges are more likely to be fragmented).

-   `p` passability of fragmented edges (probability)

-   `n_barrier` number of barriers

The arguments `p` and `n_barrier` define the probability of traversing across an edge with a barrier and the number of total barriers , respectively. Passability can be different among barriers; in this case, the user must specify the passabilities of individual barriers separately (as a vector with length `n_barrier`). The barrier effect is cumulative. For example, if one passes two barriers A (passability = $p_A$) and B (passability $p_B$) when moving from a given node node to another node, the original movement probability $m$ will be $m p_Ap_B$.

Just like `ptsource()`, this function will take an output from `brnet()` function but also accepts any adjacency matrix. `ggbrnet()` will help you visualize fragmented edges by specifying `edge_weight = "passability"`.

Example: `random` pattern:

```{r frgm1}
brnet(n_patch = 30) %>%
  frgm(rate = 0.1,
       p = 0.1,
       n_barrier = 5,
       pattern = "random") %>%
  ggbrnet(edge_weight = "passability")
```

Example: `upstream` pattern:

```{r frgm3}
brnet(n_patch = 30) %>%
  frgm(rate = 0.1,
       p = 0.1,
       n_barrier = 5,
       pattern = "upstream") %>%
  ggbrnet(edge_weight = "passability")
```

Example: `upstream` pattern:

```{r frgm4}
brnet(n_patch = 30) %>%
  frgm(rate = 0.1,
       p = 0.1,
       n_barrier = 5,
       pattern = "downstream") %>%
  ggbrnet(edge_weight = "passability")
```
