---
title: "ggbrnet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggbrnet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(mcbrnet)
```

## Basic usage

`ggbrnet()` is a wrapper of `ggraph` functions for easy visualization of a network produced by `brnet()` .

## Quick start

`ggbrnet()` takes return values of `brnet()`, `adjacency_matrix` and `df_patch`. By default, it colors patches based on environmental values `df_patch$environment`:

```{r}
x <- brnet(plot = FALSE)
ggbrnet(adjacency_matrix = x$adjacency_matrix,
        patch_attr = x$df_patch)
```

## Patch color

**Arguments:** `patch_color`

Users can change how they color patches by specifying `patch_color`. Default is `env` referring to the environment column in `patch_attr` argument. Another option is `disturb` referring to the disturbance column:

```{r, include=F}
set.seed(1)
```

```{r}
# patch ID
ggbrnet(adjacency_matrix = x$adjacency_matrix,
        patch_attr = x$df_patch,
        patch_color = "disturb")
```

If `patch_color` is neither `env` or `disturb`, the function takes it as an ordinary color name, for example:

```{r, include=F}
set.seed(1)
```

```{r}
# patch ID
ggbrnet(adjacency_matrix = x$adjacency_matrix,
        patch_attr = x$df_patch,
        patch_color = "salmon")
```

## Patch label

**Arguments:** `patch_label`

Users may add patch labels using the argument `patch_label`:

```{r, include=F}
set.seed(1)
```

```{r}
# patch ID
ggbrnet(adjacency_matrix = x$adjacency_matrix,
        patch_attr = x$df_patch,
        patch_label = "patch")
```

```{r, include=F}
set.seed(1)
```

```{r}
# branch ID
ggbrnet(adjacency_matrix = x$adjacency_matrix,
        patch_attr = x$df_patch,
        patch_label = "branch")
```

```{r, include=F}
set.seed(1)
```

```{r}
# number of upstream contributing patches
ggbrnet(adjacency_matrix = x$adjacency_matrix,
        patch_attr = x$df_patch,
        patch_label = "n_upstream")
```

Users may specify additional arguments passed to `ggraph::geom_node_label()` to tweak details, for example:

```{r}
# branch ID adjust label position
ggbrnet(adjacency_matrix = x$adjacency_matrix,
        patch_attr = x$df_patch,
        patch_label = "branch",
        nudge_x = 0.2)
```

## Patch size

To change patch size, specify `patch_size`:

```{r, include=F}
set.seed(1)
```

```{r}
ggbrnet(adjacency_matrix = x$adjacency_matrix,
        patch_attr = x$df_patch,
        patch_size = 1)
```
