---
title: "igpsim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{igpsim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(mcbrnet)
```

## Basic usage

`igpsim()` simulates tri-trophic food web dynamics with intraguild predation in space. The function employs a discrete time-series model (an extension of the Nicholson-Bailey model), which is detailed in [Pomeranz et al. 2023](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.4648). The food web dynamics are simulated through (1) local predator-prey interactions within a habitat patch, (2) immigration, and (3) emigration.

The function returns:

-   `df_dynamics` data frame containing simulated food web dynamics\*.

    -   `timestep`: time-step.
    -   `patch_id`: patch ID.
    -   `carrying_capacity`: carrying capacity at each patch.
    -   `disturbance`: disturbance-induced mortality at patch x and time-step t.
    -   `species`: species ID.
    -   `abundance`: abundance of species i at patch x.
    -   `fcl`: food chain length

-   `df_species` data frame containing species attributes.

    -   `species`: species ID.
    -   `mean_abundance`: mean abundance (arithmetic) of species i across sites and time-steps.
    -   `p_dispersal`: dispersal probability of species i.

-   `df_patch` data frame containing patch attributes.

    -   `patch_id`: patch ID.
    -   `fcl`: temporal average of food chain length.
    -   `carrying_capacity`: carrying capacity at each patch.
    -   `disturbance`: disturbance-induced mortality at each patch.

-   `df_int` data frame containing interaction parameters.

    -   `interaction`: column identifying interaction pairs
    -   `conv_eff`: conversion efficiency
    -   `attak_rate`: attack rate
    -   `handling_time`: handling time

-   `df_xy_coord` xy coordinates for habitat patches (`NULL` when `distance_matrix` or `dispersal_matrix` is provided)

-   `distance_matrix` distance matrix used in the simulation.

## Quick start

The following script simulates tri-trophic dynamics with `n_patch = 5`, which assumes five habitat patches randomly distributed over a square space. By default, `igpsim()` simulates food web dynamics with 200 warm-up (initialization with species introductions: `n_warmup`), 200 burn-in (burn-in period with no species introductions: `n_burnin`), and 1000 time-steps for records (`n_timestep`).

```{r, results='hide', message=F}
igp <- igpsim(n_patch = 5)
```

As in `mcsim()`, the simulated dynamics can be visualized by `plot = TRUE`, which will show five sample patches:

```{r, message=F, fig.width=11, results='hide'}
igp <- igpsim(n_patch = 5, plot = TRUE)
```

A named list of return values:

```{r}
igp
```

## Custom: `brnet()` + `igpsim()`

`brnet()` outputs are compatible with `igpsim()`. For example, `.$distance_matrix` may be used to inform arguments in `igpsim()`. By providing the distance matrix, the following script will simulate food web dynamics in a random branching network produced by `brnet()` function:

```{r, results='hide', message = F, fig.width=11}
patch <- 100

net <- brnet(n_patch = patch,
             p_branch = 0.5)

igp <- with(net,
            igpsim(n_patch = patch, 
                   distance_matrix = distance_matrix,
                   plot = TRUE)
            )
```

## Custom: parameter detail

Users can tweak (1) food web attributes, (2) patch attributes, and (3) landscape structure.

### Food web attributes

**Arguments:** `r_b`, `conv_eff`, `attack_rate`, `handling_time`, `s`

Food web attributes are determined based on the maximum reproductive rate of the basal species (`r_b`), conversion efficiency `conv_eff`, attack rate `attack_rate`, handling time `handling_time`, and switching parameter `s`.

*Basal species* -- `r_b` is one of the parameters defining the population growth of the basal species, modeled as follows:

$$
B_{t} = \frac{B_{t-1}r_b}{1 + \frac{r_b - 1}{K}B_{t-1}},
$$

where $B_t$ is the abundance of the basal species at time $t$, $r_b$ is the maximum growth rate (= `r_b`), and $K$ is the carrying capacity (= `carrying_capacity`; see **Patch attributes**). The parameters $r_b$ and $K$ may vary by habitat; to model such variations, users may supply vectors of `r_b` and `carrying_capacity`, whose length is equal to `n_patch`. The function assumes these values are supplied in order of patch 1, 2, 3, ..., `n_patch`. Thus, care must be taken to match this order with those in, e.g., `distance_matrix`.

*Consumers (intraguild prey and predator)* -- The predator-prey interactions are modeled with the Nicholson-Bailey model, which was extended to account for intraguild predation (see [Pomeranz et al. 2023, Ecosphere; equations 5 -- 9](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.4648) for details). The function assumes the discrete version of the Holling's Type-II functional response, in which attack rate (`atttack_rate`) and handling time (`handling_time`) define the survival function $f(\cdot)$ of the prey as follows:

$$
\begin{aligned}
f(B, C) &= \exp(-\frac{a_{BC}C}{1 + a_{BC}h_{BC}B}) &&\text{C on B},\\
f(B, P) &= \exp(-\frac{(1 + \phi)a_{BP}P}{1 + a_{BP}h_{BP}B}) &&\text{P on B},\\
f(C, P) &= \exp(-\frac{(1 - \phi)a_{CP}P}{1 + a_{CP}h_{CP}C}) &&\text{P on C},\\
\end{aligned}
$$

where $C$ and $P$ are the abundances of intraguild prey and predator, respectively, $a_{ij}$ the attack rate of consumer $j$ on prey $i$ (= `attack_rate`), and $h_{ij}$ the handling time (= `handling_time`). If these arguments are supplied as scalars, then the function assumes the constant values for all the interactions (i.e., $a_{BC} = a_{BP} = a_{CP} = const.$ and $h_{BC} = h_{BP} = h_{CP} = const.$). Where appropriate, users may apply different values to these by supplying vectors; in this case, the function assumes the parameter values appear in order of $a_{BC}$ ($h_{BC}$), $a_{BP}$ ($h_{BC}$), and $a_{CP}$ ($h_{CP}$). The parameter $\phi$ quantifies the switching tendency of the intraguild predator, defined as follows:

$$
\phi = \frac{s(B - C)}{B+C}
$$

The parameter $s$ (= `s`) determines the likelihood of switching to the basal species, with higher values of $s$ indicating the greater switching tendency to the basal species. Lastly, conversion efficiency (= `conv_eff`) quantifies the effectiveness of consumers transforming the capture prey into consumer's abundance. `conv_eff` must be a scalar (supply identical values to all the interactions) or must be supplied in order of $BC$, $BP$, and $CP$.

Note that the function assumes the following order of ecological events: $B$'s reproduction $\rightarrow$ $C$'s predation on $B$ $\rightarrow$ $C$'s reproduction $\rightarrow$ $P$'s predation on $B$ and $C$ $\rightarrow$ $P$'s reproduction.

### Patch attributes

**Arguments:** `carrying_capacity` `p_disturb`, `i_disturb`, `phi_disturb`

Patch attributes are characterized by carrying capacity (= `carrying_capacity`) and disturbance (probability of occurrence = `p_disturb`, disturbance intensity = `i_disturb`, temporal precision of disturbance intensity = `phi_disturb`). `carrying_capacity` and `i_disturb` can be supplied as patch specific values, provided that the vector lengths match `n_patch`. The other arguments `p_disturb` and `phi_disturb` should be given as scalars. `p_disturb` controls how often disturbance occurs, while `phi_disturb` regulates the temporal variability of the disturbance intensity when it occurs (greater values of `phi_disturb` indicating ***less*** temporal variability).

Note that disturbance events in this function is designed to resemble regional disturbance, which affects all habitat patches when it occurs. Thus, disturbance heterogeneity within a landscape should be introduced through `i_disturbance` argument if desired.

### Landscape attributes

**Arguments:** `xy_coord`, `distance_matrix`, `landscape_size`, `theta`

See **Article** for `mcsim()`

### Others

**Arguments:** `n_warmup`, `n_burnin`, `n_timestep`

See **Article** for `mcsim()`

## Model description

Full model descriptions are available at [Pomeranz et al. 2023, Ecosphere](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.4648).
