---
title: "mcsim: metacommunity simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mcsim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(mcbrnet)
```

## Basic usage

This function simulates metacommunity dynamics. Community dynamics are modeled based on the Beverton-Holt function. Although this function is designed to be compatible with `brnet()`, users can provide their distance matrix to simulate dynamics in any landscape. The key arguments are the number of habitat patches (`n_patch`) and the number of species in a metacommunity (`n_species`). Metacommunity dynamics are simulated through (1) local dynamics (population growth and competition among species), (2) immigration, and (3) emigration.

Sample script:

```{r, eval = F}
library(mcbrnet)
mc <- mcsim(n_patch = 5, n_species = 5)
```

The function returns:

-   `df_dynamics` a data frame containing simulated metacommunity dynamics\*.

    -   *timestep*: time-step.
    -   *patch_id*: patch ID.
    -   *mean_env*: mean environmental condition at each patch.
    -   *env*: environmental condition at patch x and time-step t.
    -   *carrying_capacity*: carrying capacity at each patch.
    -   *species*: species ID.
    -   *niche_optim*: optimal environmental value for species i.
    -   *r_xt*: reproductive number of species i at patch x and time-step t.
    -   *abundance*: abundance of species i at patch x.

-   `df_species` a data frame containing species attributes.

    -   *species*: species ID.
    -   *mean_abundance*: mean abundance (arithmetic) of species i across sites and time-steps.
    -   *r0*: maximum reproductive number of species i.
    -   *niche_optim*: optimal environmental value for species i.
    -   *sd_niche_width*: niche width for species i.
    -   *p_dispersal*: dispersal probability of species i.

-   `df_patch` a data frame containing patch attributes.

    -   *patch*: patch ID.
    -   *alpha_div*: alpha diversity averaged across time-steps.
    -   *mean_env*: mean environmental condition at each patch.
    -   *carrying_capacity*: carrying capacity at each patch.
    -   *connectivity*: functional connectivity at each patch.

-   `df_diversity` a data frame containing diversity metrics (α, β, and γ).

-   `distance_matrix` a distance matrix used in the simulation.

-   `interaction_matrix` a species interaction matrix, in which species X (column) influences species Y (row).

\*NOTE: The warm-up and burn-in periods will not be included in return values.

## Quick start

The following script simulates metacommunity dynamics with `n_patch = 5` and `n_species = 5`. By default, `mcsim()` simulates metacommunity dynamics with 200 warm-up (initialization with species introductions: `n_warmup`), 200 burn-in (burn-in period with no species introductions: `n_burnin`), and 1000 time-steps for records (`n_timestep`).

```{r, results='hide', message=F}
mc <- mcsim(n_patch = 5, n_species = 5)
```

Users can visualize the simulated dynamics using `plot = TRUE`, which will show five sample patches and species that are randomly chosen:

```{r, message=F, fig.width=11, results='hide'}
mc <- mcsim(n_patch = 5, n_species = 5, plot = TRUE)
```

A named list of return values:

```{r}
mc
```

## Customize: `brnet()` + `mcsim()`

Return values of `brnet()` are compatible with `mcsim()`. For example, `df_patch$environment`, `df_patch$n_patch_upstream`, and `df_patch$distance_matrix` may be used to inform parameters of `mcsim()`:

```{r, results='hide', message = F, fig.width=11}
patch <- 100
net <- brnet(n_patch = patch, p_branch = 0.5, plot = F)
mc <- mcsim(n_patch = patch, n_species = 5,
            mean_env = net$df_patch$environment,
            carrying_capacity = net$df_patch$n_patch_upstream*10,
            distance_matrix = net$distance_matrix,
            plot = T)

```

## Customize: detailed parameters

Users may use the following arguments to custom metacommunity simulations regarding (1) species attributes, (2) competition, (3) patch attributes, and (4) landscape structure.

### Species attributes

**Arguments:** `r0`, `niche_optim` OR `min_optim` and `max_optim`, `sd_niche_width` OR `min_niche_width` and `max_niche_width`, `niche_cost`, `p_dispersal`

Species attributes are determined based on the maximum reproductive rate `r0`, optimal environmental value `niche_optim` (or `min_optim` and `max_optim` for random generation of `niche_optim`), niche width `sd_niche_width` (or `min_niche_width` and `max_niche_width` for random generation of `sd_niche_width`) and dispersal probability `p_dispersal` (see **Model description** for details).

For optimal environmental values (niche optimum), the function by default assigns random values to species as: $\mu_i \sim Unif(\mu_{min}, \mu_{max})$, where users can set values of $\mu_{min}$ and $\mu_{max}$ using `min_optim` and `max_optim` arguments (default: `min_optim = -1` and `max_optim = 1`). Alternatively, users may specify species niche optimums using the argument `niche_optim` (scalar or vector). If a single value or a vector of `niche_optim` is provided, the function ignores `min_optim` and `max_optim` arguments.

Similarly, the function by default assigns random values of $\sigma_{niche}$ to species as: $\sigma_{niche,i} \sim Unif(\sigma_{niche,min}, \sigma_{niche,max})$. Users can set values of $\sigma_{niche,min}$ and $\sigma_{niche,max}$ using `min_niche_width` and `max_niche_width` arguments (default: `min_niche_width = 0.1` and `max_niche_width = 1`). If a single value or a vector of `sd_niche_width` is provided, the function ignores `min_niche_width` and `max_niche_width` arguments.

The argument `niche_cost` determines the cost of having wider niche. Smaller values imply greater costs of wider niche (i.e., decreased maximum reproductive rate; default: `niche_cost = 1`). To disable (no cost of wide niche), set `niche_cost = Inf`.

For other parameters, users may specify species attributes by giving a scalar (assume identical among species) or a vector of values whose length must be one or equal to `n_species`. Default values are `r0 = 4`, `sd_niche_width = 1`, and `p_dispersal = 0.1`.

### Competition

**Arguments:** `interaction_type`, `alpha` OR `min_alpha` and `max_alpha`

The argument `interaction_type` determines whether interaction coefficient `alpha` is a constant or random variable. If `interaction_type = "constant"`, then the interaction coefficients $\alpha_{ij, i \ne j}$ for any pairs of species will be set as a constant `alpha` (i.e., off-diagonal elements of the interaction matrix). If `interaction_type = "random"`, $\alpha_{ij, i \ne j}$ will be drawn from a uniform distribution as $\alpha_{ij, i \ne j} \sim Unif(\alpha_{min}, \alpha_{max})$ with corresponding arguments `min_alpha` and `max_alpha`. The argument `alpha` is ignored under the scenario of random interaction strength (i.e., `interaction_type = "random"`). Note that the diagonal elements of the interaction matrix ($\alpha_{ii}$) are always 1.0 regardless of `interaction_type`, as `alpha` is the strength of interspecific competition relative to that of intraspecific competition (see **Model description**). By default, `interaction_type = "constant"` and `alpha = 0`.

### Patch attributes

**Arguments:** `carrying_capacity`, `mean_env`, `sd_env`, `spatial_auto_cor`, `phi`

The arguments `carrying_capacity` (default: `carrying_capacity = 100`) and `mean_env` (default: `mean_env = 0`) determines mean attributes of habitat patches, which can be a scalar (assume identical among patches) or a vector (length must be equal to `n_patch`).

The arguments `sd_env` (default: `sd_env = 0.1`), `spatial_auto_cor` (default: `spatial_auto_cor = FALSE`) and `phi` (default: `phi = 1`) determine spatio-temporal dynamics of environmental values. `sd_env` determines the magnitude of temporal environmental fluctuations. If `spatial_auto_cor = TRUE`, the function models spatial autocorrelation of temporal environmental fluctuation based on a multi-variate normal distribution. The degree of spatial autocorrelation would be determined by `phi`, the parameter describing the strength of distance decay in spatial autocorrelation (see **Model description**).

### Landscape structure

**Arguments:** `xy_coord` OR `distance_matrix`, `landscape_size`, `theta`

These arguments define landscape structure. By default, the function produces a square-shaped landscape (`landscape_size = 10` on a side) in which habitat patches are distributed randomly through a Poisson point process (i.e., x- and y-coordinates of patches are drawn from a uniform distribution). The parameter θ describes the shape of distance decay in species dispersal (see **Model description**) and determines patches' structural connectivity (default: `theta = 1`). Users can define their landscape by providing either `xy_coord` or `distance_matrix` (`landscape_size` will be ignored if either of these arguments is provided). If `xy_coord` is provided (2-column data frame denoting x- and y-coordinates of patches, respectively; `NULL` by default), the function calculates the distance between patches based on coordinates. Alternatively, users may provide `distance_matrix` (the object must be `matrix`), which describes the distance between habitat patches. The argument `distance_matrix` overrides `xy_coord`.

### Others

**Arguments:** `n_warmup`, `n_burnin`, `n_timestep`

The argument `n_warmup` is the period during which species introductions occur (default: `n_warmup = 200`). The initial number of individuals introduced follows a Poisson distribution with a mean of 0.5 and independent across space and time. This random introduction events occur multiple times over the `n_warmup` period, in which `propagule_interval` determines the timestep interval of the random introductions (default: `propagule_interval = ceiling(n_warmup / 10)`).

The argument `n_burnin` is the period that will be discarded as *burn-in* to remove the influence of initial values (default: `n_burnin = 200`). During the burn-in period, species introductions do not occur.

The argument `n_timestep` is the simulation peiord that is recorded in the return `df_dynamics` (default: `n_timestep = 1000`). As a result, with the default setting, the function simulates 1400 timesteps (`n_warmup` + `n_burnin` + `n_timestep` = 1400) but returns only the last 1000 timesteps as the resulting metacommunity dynamics. All the derived statistics (e.g., diversity metrics in `df_diversity` and `df_patch`) will be calculated based on the results during `n_timestep`.

## Model description

Full model descriptions are available at: <https://doi.org/10.1073/pnas.2105574118>
